A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-04-23 papers/water-SAFT/paper.tex
Made a few changes to paper.  Moved paper directory to papers/water-SAFT. Tested everything on quipu.

Water stands out among liquids for its extraordinary properties.
Water is the only natural substance that is found in solid, liquid and
gaseous states at temperatures and pressures normally found on Earth.
Water has a density maximum at 4$^\circ$C, and expands upon freezing.
Water has the highest heat capacity, latent heat of fusion and
vaporization, thermal conductivity and surface tension of any
non-metallic liquid;  water also has an unusually high dielectric constant.
Much of the unusual behavior of water is explained by the ability of
each water molecule to form \emph{four} strong, directional hydrogen
bonds---two acceptors and two donors.  This directional bonding
dictates the open structure of ice, leading to thermal expansion upon
freezing.

@@

\needsworknow{Add discussion and citations for more recent papers}

@@

\section{Theory and Methods}
We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water, unlike simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations, requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}.  \needsworknow{Expand this a
  bit to accurately describe SAFT and its advantages.}

@@

The association functional we use is a modified version of that of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}.\footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.} 
which includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$. This, however, is
based on the SAFT-HS equation of state, rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR} which is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}.  The modification to use the SAFT-VR
equation of state involves adding a correction term, see
Equation~\ref{eq:gSW}.  \fixme{needs work}

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3~nm, 0.5~nm, and 1~nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.
\needsworknow{Expand or add to this description of density profiles and
oscillations for single rod?}

@@

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of radii 0.3~-~1.0~nm at separation $d$ 
(see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to $d=1.3$~nm. At small
separations there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. 
Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. This critical separation for the transition to liquid depends
on the radii of the rods, and is about 0.54~nm for the rods shown in 
Figure~\ref{fig:density-rods}. 
Since we are looking at only hard-walled
non-interacting rods, the critical separation will be
different for any system where there is an appreciable direct force between rods. 
The density oscillations seen around the rods in Figure~\ref{fig:density-rods} 
are very similar to
those for the single hydrophobic rod in water (see Figure~\ref{fig:density-single-rod}).
At small separations, the oscillations around the two rods look as if the two rods were one
solid ``stadium''-shaped object (a rectangle with semi-circles on the ends).

@@

To explain this, we consider a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for larger 
rods ... \needsworknow{Explain ``bowing'' out of vapor area for larger rods.}

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use this model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown graphically in 
Figure~\ref{fig:rods-energy-vs-distance}, and they match fairly well for the 
0.5~-~1.0~nm rods with differences between 1.3~-~5.5\%. We hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction
well(32\% difference).

@@

We simulate four hydrophobic rods in water using the same radii as our
two rods ($r = 0.3$~nm~-~$1.0$~nm)and
separation distances between $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior as the two rods
where vapor is present in the space between the rods until a critical separation
where it becomes filled with water (shown in Fig. \ref{fig:density-four-rods}). 
For large $d$,
the rods behave like separate single rods.  Free energy per length
is plotted against rod separation in Figure \ref{fig:four-rods-energy-vs-distance}.
From the slope we calculate the force per length to be [number].

The rounded square formed by the area of the four rods and the space
inbetween them has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of
only the four rods, which results in a critical separation of
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r. \label{criticalfourrods}
\end{equation}
We see in Figure \ref{fig:four-rods-energy-vs-distance} that the $r = 3.0$~nm
rod has the largest difference between simulation and prediction with
a 23\% difference, and better agreement for larger radii (between 0.5\% and 5.6\%).

@@

Figure~\ref{fig:density-sphere} shows the density profile for several
hard sphere radii, plotted together with the results of the same
SPC/E molecular dynamics simulation shown in
Figure~\ref{fig:sphere-energy-vs-diameter}\cite{huang2001shs}.  The
agreement with simulation is quite reasonable.  The largest
disagreement involves the density at contact, which according to the
contact value theorem cannot agree, since the free energies do not
agree.c

@@

The free energy per area for water surrounding a hydrophobic rod shows the
expected behavior with an asymptote equal to the surface tension.
Based on this result, we expect to be
able to classically describe rods with a radius of about 0.5~nm or larger. 
For two hydrophobic rods surrounded by water, we see a transition from a vapor-filled
stadium shape to two rods with liquid between them. A
classical treatment of the critical separation for this transition
works well for rods larger than 0.5~nm. Further evidence for the
accuracy of this classical treatment is the calculation of the force on the
stadium shape. The force per length is constant and equal to twice the surface tension.

>>>>

Water stands out among liquids for its extraordinary properties.
Water is the only natural substance that is found in solid, liquid and
gaseous states at temperatures and pressures normally found on Earth.
Water has a density maximum at 4$^\circ$C, and expands upon freezing.
Water has the highest heat capacity, latent heat of fusion and
vaporization, thermal conductivity and surface tension of any
non-metallic liquid.  Water also has an unusually high dielectric constant.
Much of the unusual behavior of water is explained by the ability of
each water molecule to form \emph{four} strong, directional hydrogen
bonds---two acceptors and two donors.  This directional bonding
dictates the open structure of ice, leading to thermal expansion upon
freezing.

@@

\needsworknow{Add discussion and citations for more recent papers}
Test

@@

\section{Theory and Methods}
We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water is unlike simple
fluids (which are mainly defined by van der Waals forces and weak electrostatic
interations) because it requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}.  \needsworknow{Expand this a
  bit to accurately describe SAFT and its advantages.}

@@

The association functional we use is a modified version of that of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating} \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.} 
which includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$. This, however, is
based on the SAFT-HS equation of state, rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR} which is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}.  The modification to use the SAFT-VR
equation of state involves adding a correction term, see
Equation~\ref{eq:gSW}.  \fixme{needs work}

@@

We move into two dimensions by studying a single hydrophobic rod
immersed in water. We use a lattice spacing of about 0.01~nm (0.2 bohrs), 
which is a reasonable resolution based on several lattice spacing calculations.
The results for this spacing of 0.01~nm are within 0.5\% of the free energy 
and a few percent of the contact
density for the highest resolution tested.
Figure~\ref{fig:density-single-rod} shows density
profiles for three different radii (0.3~nm, 0.5~nm, and 1~nm) which have the
expected oscillations outside the rod that reduce to the bulk saturated liquid
density as the distance away from the rods increases. The small radius
rod has comparatively larger oscillations and a higher contact density.
\needsworknow{Expand or add to this description of density profiles and
oscillations for single rod?}

@@

We now look at the more interesting problem of two rods in water. We continue to
use hydrophobic rods of radii 0.3~-~1.0~nm at separation $d$ 
(see Figure~\ref{fig:rods}), ranging from $d=0$ 
(touching rods) to $d=1.3$~nm. At small
separations there is only vapor between the rods. As the rods are
pulled apart, the vapor region expands until some critical separation is
reached and liquid water fills the region between the rods. 
Figure~\ref{fig:density-rods} 
shows density profiles before and after this transition
for rods of radius 0.5~nm. This critical separation for the transition to liquid depends
on the radii of the rods, which is about 0.54~nm for the rods shown in 
Figure~\ref{fig:density-rods}.
Since we are looking at only hard-walled
non-interacting rods, the critical separation will be
different for any system where there is an appreciable direct force between rods. 
The density oscillations seen around the rods in Figure~\ref{fig:density-rods} 
are very similar to
those for the single hydrophobic rod in water (see Figure~\ref{fig:density-single-rod}).
At small separations, the oscillations around the two rods look as if the two rods were one
solid ``stadium''-shaped object (a rectangle with semi-circles on the ends).

@@

To explain this, we consider a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for those 
rods, the water on the sides of the 'stadium' configuration will bow inward 
between the rods and the densty will still reduce to vapor where the rods 
are closest to each other.

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use this model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown graphically in 
Figure~\ref{fig:rods-energy-vs-distance}, and they match fairly well for the 
0.5~-~1.0~nm rods with differences between 1.3~-~5.5\%. We hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction
well (32\% difference).

@@

We simulate four hydrophobic rods in water using the same radii as our
two rods ($r = 0.3$~nm~-~$1.0$~nm) and
separation distances between $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior as the two rods
where vapor is present in the space between the rods until a critical separation
where it becomes filled with water and the rods behave like separate single rods
(shown in Fig. \ref{fig:density-four-rods}).  Free energy per length
is plotted against rod separation in Figure \ref{fig:four-rods-energy-vs-distance} and
from the slope we calculate the force per length to be [number].

The 'rounded square' formed by the area of the four rods and the space
inbetween them has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of
only the four rods, which results in a critical separation of
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r. \label{criticalfourrods}
\end{equation}
We see in Figure \ref{fig:four-rods-energy-vs-distance} that the $r = 0.3$~nm
rod has the largest difference between simulation and prediction with
a 23\% difference, and better agreement for larger radii rods (between 0.5\% and 5.6\%).

@@

Figure~\ref{fig:density-sphere} shows the density profile for several
hard sphere radii, plotted together with the results of the same
SPC/E molecular dynamics simulation shown in
Figure~\ref{fig:sphere-energy-vs-diameter}\cite{huang2001shs}.  The
agreement with simulation is quite reasonable.  The largest
disagreement involves the density at contact, which according to the
contact value theorem cannot agree, since the free energies do not
agree.

@@

The free energy per area for water surrounding a hydrophobic rod shows the
expected behavior with an asymptote equal to the surface tension.
Based on this result, we expect to be
able to classically describe rods with a radius of about 0.5~nm or larger. 
For two hydrophobic rods surrounded by water, we see a transition from a vapor-filled
stadium shape to two rods with liquid between them. A
classical treatment of the critical separation for this transition
works well for rods larger than 0.5~nm.  Further evidence for the
accuracy of this classical treatment is the calculation of the force on the
stadium shape. The force per length is constant and equal to twice the
surface tension. A simulations of hydrophobic spheres agrees with
SPC/E calculations and a system with four hydrophobic rods exhibited
similar behavior as the two rod.
